---
title: "A comprehensive review of proteomics signature discovery for Tuberculosis"
author: "Zaynab Mousavian"
output: html_document
date: "2023-05-01"
---

### Loading R packages

```{r}
library(readxl)
library(ggplot2)
library(RColorBrewer)
library(patchwork)
```

### The function for searching uniprot IDs in the Uniprot databse

```{r}
uniprotSearch <- function(proteinName){
  proteinName <- gsub("-","",proteinName)
  line <- which(toupper(trimws(proteinName))==uniprot_db$`Entry Name`)
  if (length(line)==0) line <- grep(paste0("\\b",trimws(proteinName),"\\b"), uniprot_db$`Gene Names`,ignore.case = TRUE)
  uniprotID <- uniprot_db[line,"Entry"]
  if (length(line)==0) uniprotID="NULL"
  return(uniprotID)
}
```

### Mapping published signatures to uniprot IDs (No need to run)

```{r}
convertFunction <- function(inputString) {
  inputValues <- strsplit(inputString, ",")[[1]]
  outputValues <- lapply(inputValues, function(y) uniprotSearch(y))
  outputString <- paste(unlist(outputValues),collapse = ",")
  return(outputString)
}
uniprot_db <- read_excel("../uniprot-human-DB.xlsx")
uniprot_db$`Entry Name` <- unlist(lapply(uniprot_db$`Entry Name`, function(x) strsplit(x,"_")[[1]][1]))
file <- read.delim("../Final version of summary file.csv",sep=";")
names(file)=file[1,]
file=file[-1,]
signatures_list <- file$`Signature (only proteins)`
Uniprot_list <- unlist(lapply(signatures_list,function(y) convertFunction(y)))
split_list1 <- strsplit(signatures_list, ",")
split_list2 <- strsplit(Uniprot_list, ",")

# Merge the corresponding elements of list1 and list2 into one list
merged_list <- mapply(paste, split_list1, split_list2)

max_length <- max(sapply(merged_list, length))

# Pad the shorter lists with NAs to make them the same length as the longest list
for (i in seq_along(merged_list)) {
  if (length(merged_list[[i]]) < max_length) {
    merged_list[[i]] <- c(merged_list[[i]], rep(NA, max_length - length(merged_list[[i]])))
  }
}

merged_df <- as.data.frame(do.call(rbind, merged_list))
merged_df <- cbind(paste(file$Study,file$Year,sep=","),merged_df)
write.csv(merged_df,file="Converted_signatures.csv")
```

### Mapping protein annotation from the Human Protein Atlas database to uniprot IDs (No need to run)

```{r}
proteinAnnotation = read.delim("ProteinMapping.txt",sep=" ")
row.names(proteinAnnotation) = proteinAnnotation$display_name

list1=lapply(proteinAnnotation$display_name,function(x) uniprotSearch(x))
proteinAnnotation$UniprotID <- list1
proteinAnnotation$UniprotID <- lapply(proteinAnnotation$UniprotID,function(x) as.character(x))
proteinAnnotationData <- apply(proteinAnnotation,2,as.character)
write.csv(proteinAnnotationData,file="proteinMappingWithUniprotID.csv",quote=FALSE)
```

### Mapping KEGG signaling pathways from the MSigDB database to uniprot IDs (No need to run)

```{r}
KEGG <- read.csv("c2.cp.kegg.v2023.1.Hs.symbols.csv",header = FALSE,sep=";")

uniprotSearch <- function(proteinName){
  proteinName <- gsub("-","",proteinName)
  line <- which(toupper(trimws(proteinName))==uniprot_db$`Entry Name`)
  if (length(line)==0) line <- grep(paste0("\\b",trimws(proteinName),"\\b"), uniprot_db$`Gene Names`,ignore.case = TRUE)
  uniprotID <- uniprot_db[line,"Entry"]
  if (length(line)==0) uniprotID="NULL"
  return(uniprotID)
}

KEGG_converted = matrix(NA, nrow = nrow(KEGG), ncol = ncol(KEGG)-2)
for (i in 1:nrow(KEGG_converted))
  for (j in 1:ncol(KEGG_converted))
    if (KEGG[i,j+2]!="")
       KEGG_converted[i,j] <- as.character(uniprotSearch(KEGG[i,j+2]))
    
save(KEGG_converted,file="KEGG_converted.RData")
```

### Extracting uniprot IDs for all published signatures

```{r}
# remove HUMAN_ from uniprot Entry name
uniprot_db <- read_excel("../uniprot-human-DB.xlsx")
uniprot_db$`Entry Name` <- unlist(lapply(uniprot_db$`Entry Name`, function(x) strsplit(x,"_")[[1]][1]))

# read signatures from all publications (86 studies) which have been already converted to Uniprot IDs 
file <- read.csv("Converted_signatures_final.csv",sep=";")
file <- file[,-c(1,2)]
file <- data.frame(lapply(file, as.character), stringsAsFactors = FALSE)

# read the information of all publications
publications <- read.delim("../Final version of summary file.csv",sep=";")
names(publications)=publications[1,]
publications=publications[-1,]

# extract uniprot IDs from all publications (183 unique uniprotIds which two values are "NULL" and NA -> 181 uniprotIds)
my_fun <- function(x) {
    sapply(strsplit(trimws(x), "\\s+"), "[", 2)
}
uniprot_signatures <- data.frame(apply(file, 2, my_fun))
sort(table(unlist(uniprot_signatures)))
```

### Visualize number of publications over years (Figure 1:A)

```{r}
Num_df <- publications[,"Year",drop=FALSE]
Num_df$Value=1
Num_df <- aggregate(Value~Year,data=Num_df,FUN = sum)
Num_df <- Num_df[-which(Num_df$Year==2023),]

publications_plot <- ggplot(Num_df,aes(x=Year,y=Value))+
  geom_bar(stat="identity",width=0.5,position=position_dodge(0))+
  theme_bw()+
  theme(axis.title = element_text(size=7),axis.text = element_text(size=5))+
  labs(x="Year",y="Number of publications")
  

pdf(file="Figures/Publications-Year.pdf",width=3,height = 2)
publications_plot
dev.off()
  
```

### Finding association between proteins and KEGG signaling pathways

```{r}
# read all KEGG pathways (186 pathways)
KEGG <- read.csv("c2.cp.kegg.v2023.1.Hs.symbols.csv",header = FALSE,sep=";")
# read all KEGG pathways which have been already converted to uniprotIds (186 pathways)
load("KEGG_converted.RData")
KEGG_converted <- cbind(KEGG$V1,KEGG_converted)

# finding association between signatures and KEGG pathways by using uniprotIds (97 uniprotIds (all have entry name) are associated with at least one pathway)
KEGG_results=data.frame()
for (i in 1:nrow(uniprot_signatures))
  for (j in 1:ncol(uniprot_signatures))
    if (!is.na(uniprot_signatures[i,j])&!is.null(uniprot_signatures[i,j])&uniprot_signatures[i,j]!="NULL")
    {
      uniprotID=uniprot_signatures[i,j]
      lines <- apply(KEGG_converted,1, function(x) length(grep(uniprotID,x))>0)
      if (length(which(lines))!=0)
      {
        temp=data.frame(Pathway=KEGG_converted[lines,1],ProteinID=uniprotID)
        KEGG_results <<- rbind(temp,KEGG_results)
      }
    }

KEGG_results = KEGG_results[!duplicated(KEGG_results), ]
KEGG_results = KEGG_results[order(KEGG_results$Pathway),]
# sort(table(KEGG_results$Pathway))

KEGG_results$Pathway=gsub("KEGG_","",KEGG_results$Pathway)
KEGG_results$Pathway=gsub("_PATHWAY","",KEGG_results$Pathway)
KEGG_results$Gene <- unlist(lapply(KEGG_results$ProteinID, function(x) uniprot_db[match(x, uniprot_db$Entry),"Entry Name"]))
write.csv(KEGG_results,file="KEGG_results.csv",quote=FALSE)

# convert uniprotIds to Uniprot Entry name for all signatures (178 unique Uniprot Entry name which one value is NA -> 177 Entry names)
gene_signatures <- apply(uniprot_signatures, 2, function(x) uniprot_db[match(x, uniprot_db$Entry),"Entry Name"])
gene_signatures <- as.data.frame(gene_signatures)
number_publications <- as.data.frame(table(unlist(gene_signatures)))
row.names(number_publications) <- number_publications$Var1
```

### Calculate frequencies for visualization

```{r}
# calculate the lenght of each pathway name and each protein name
pathway_name_size <- unlist(lapply(names(sort(table(KEGG_results$Pathway))),function(x) nchar(x)))
gene_name_size <- unlist(lapply(names(sort(table(KEGG_results$Gene))),function(x) nchar(x)))

# calculate the frequency of each protein name in publications
selected_genes <- number_publications[unique(KEGG_results$Gene),]
selected_genes <- selected_genes[order(selected_genes$Freq),]
selected_genes$Var1 <- as.character(selected_genes$Var1)

# calculate the frequency of each pathway in publications
selected_pathways <- data.frame(sort(table(KEGG_results$Pathway)))
selected_pathways$Var1 <- as.character(selected_pathways$Var1)

# merge all details into one info table
info=data.frame(rbind(cbind(Name=selected_pathways$Var1,Connection=selected_pathways$Freq,Type="Pathway",Size=pathway_name_size),cbind(Name=selected_genes$Var1,Connection=selected_genes$Freq,Type="Gene",Size=gene_name_size)))
info$Connection=as.integer(info$Connection)
info$Size=as.integer(info$Size)
info=info[order(info$Type,info$Connection,info$Size),]

# extract details of pathways and proteins into two tables
pathway_info=info[info$Type=="Pathway",]
gene_info=info[info$Type=="Gene",]
pathway_order=rev(pathway_info$Name)
gene_order=rev(gene_info$Name)

# extract pathways which have connection with at least 5 proteins (80 proteins)
pathway_info <- pathway_info[which(pathway_info$Connection>4),]
pathway_order=rev(pathway_info$Name)
KEGG_results <-  KEGG_results[which(KEGG_results$Pathway%in%pathway_order),]
gene_info <-  gene_info[which(gene_info$Name%in%KEGG_results$Gene),]
gene_order=rev(gene_info$Name)

# extract remaining proteins (97 proteins)
remainingProteins = number_publications[which(!number_publications$Var1%in%gene_order),"Var1"]

# extract uniprotIds for all proteins which have name (177 proteins)
number_publications$UniprotID <- unlist(lapply(number_publications$Var1,function(x) uniprotSearch(x)))
```

### Defining color palette for protein annotation

```{r}
# read proteins annotation
proteinAnnotation = read.csv("proteinMappingWithUniprotID.csv")

number_publications$Annotation <- proteinAnnotation[match(number_publications$UniprotID,proteinAnnotation$UniprotID),"function_name"]
number_publications$Annotation <- gsub("NULL","Other",number_publications$Annotation)
number_publications$Annotation <- gsub("No annotated function","Other",number_publications$Annotation)
number_publications[which(is.na(number_publications$Annotation)),"Annotation"]="Other"
number_publications$Annotation <- factor(number_publications$Annotation,levels=c(setdiff(names(sort(table(number_publications$Annotation),decreasing = TRUE)),"Other"),"Other"))
library(RColorBrewer)
numColors <- length(unique(number_publications$Annotation)) # How many colors you need
myPalette <- c(brewer.pal(n = 8, name = "Dark2"),brewer.pal(n = 8, name = "Set1"),"darkred","darkblue","grey")
names(myPalette) <- levels(number_publications$Annotation)
myPalette["Interferon"]="turquoise"
myPalette["Hormone"]="skyblue"
myPalette["Apolipoprotein"]="#808000"
myPalette["Acute phase"]="#E41A1C"
myPalette["Enzyme inhibitor"]="black"
```

### Visualize legend for protein annotation

```{r}
# generate legend for proteins annotation
Total_annotation = number_publications$Annotation
pdf(file="Figures/guide_annotation.pdf",width=4,height =7)
ggplot(number_publications,aes(x=Var1,y=Freq,fill=Annotation))+
  geom_col()+
  scale_fill_manual(values=myPalette[Total_annotation])+
  guides(fill=guide_legend(ncol=2,byrow=TRUE))+
  labs(fill="",y="",x="")
dev.off()
```

### Visualize remaining proteins without association with pathways (Figure 1-D)

```{r}
# generate barplot for remaining proteins without connection with pathways (97 proteins)
remaining_genes <- number_publications[remainingProteins,]
remaining_genes <- remaining_genes[order(remaining_genes$Freq),]
remaining_genes$Var1 <- as.character(remaining_genes$Var1)
remaining_gene_info = data.frame(Name=remaining_genes$Var1,Connection=remaining_genes$Freq,Type="Gene")
remaining_gene_order=rev(remaining_gene_info$Name)
remaining_gene_annotation=number_publications[remaining_gene_order,"Annotation"]
remaining_gene_info$annotation <- number_publications[remaining_gene_info$Name,"Annotation"]

remaining_gene_plot <- ggplot(remaining_gene_info,aes(x=factor(Name,level=remaining_gene_order),y=Connection,fill=annotation))+
  geom_col()+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5,size=10,face=1,colour=myPalette[remaining_gene_annotation]),axis.text.y=element_text(size=9),axis.ticks.x = element_blank(),legend.position = "none")+
  labs(y="",x="",fill="")+
  scale_y_continuous(breaks=c(0,5,10,15,20),expand=c(0,0))+
  scale_fill_manual(values=myPalette[remaining_gene_annotation])
  
```

### Visualize proteins with association with pathways (Figure 1-C)

```{r}
# generate all plots for proteins which have connection with the most frequent pathways (80 proteins)
Gene_annotation=number_publications[gene_order,"Annotation"]
gene_info$annotation <- number_publications[gene_info$Name,"Annotation"]

dotplot<-ggplot(KEGG_results,aes(x=factor(Gene,level=gene_order),y=factor(Pathway,level=pathway_order)))+
  geom_point(size=3)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5,size=10,face=1,colour=myPalette[Gene_annotation]),axis.text.y=element_text(size=10),plot.margin=unit(c(0,0,0,0),"cm"))+
  labs(x="",y="")


pathway_barplot<-ggplot(pathway_info,aes(y=factor(Name,level=pathway_order),x=Connection))+
  geom_col()+
  theme_bw()+
  theme(axis.text.y=element_blank(),axis.text.x=element_text(size=10), axis.ticks.y = element_blank(), plot.margin=unit(c(0,0,-3,0),"cm"))+
  labs(x="",y="")+
  scale_x_continuous(breaks=c(0,10,20,30,40),expand=c(0,0))

gene_barplot<-ggplot(gene_info,aes(x=factor(Name,level=gene_order),y=Connection,fill=annotation))+
  geom_col()+
  theme_bw()+
  theme(axis.text.x=element_blank(),axis.text.y=element_text(size=10),axis.ticks.x = element_blank(),plot.margin=unit(c(0,-3,0,0),"cm"),legend.position = "none")+
  labs(y="",x="")+
  scale_y_continuous(breaks=c(0,5,10,15,20),expand=c(0,0))+
  scale_fill_manual(values=myPalette[Gene_annotation])

pdf("Figures/merged-plots.pdf",height = 7,width=20)
# gene_barplot+ggplot()+theme_void()+dotplot+pathway_barplot+plot_layout(widths = c(1,0.2),heights = c(0.2,1))
gene_barplot+ggplot()+theme_void()+dotplot+pathway_barplot+remaining_gene_plot+ggplot()+theme_void()+plot_layout(widths = c(1,0.055),heights = c(0.2,1,0.2))
dev.off()
```

### Calculate frequencies for showing in donut plots (Figure 1-B)

```{r}
file <- read.delim("../Final version of summary file.csv",sep=";")
names(file)=file[1,]
file=file[-1,]

statistics_data <- data.frame()
statistics_data <- rbind(statistics_data,cbind(Name="Blood/PBMC",Size=length(which(file$`Blood/PBMC`=="TRUE"))))
statistics_data <- rbind(statistics_data,cbind(Name="Serum/Plasma",Size=length(which(file$`Serum/Plasma`=="TRUE"))))
statistics_data <- rbind(statistics_data,cbind(Name="Sputum/Saliva",Size=length(which(file$`Sputum/Saliva`=="TRUE"))))
statistics_data <- rbind(statistics_data,cbind(Name="Urine",Size=length(which(file$Urine=="TRUE"))))
statistics_data <- rbind(statistics_data,cbind(Name="Other",Size=length(which(file$Other=="TRUE"))))
statistics_data$Type <- "SampleType"

temp=as.data.frame(table(file$Technology))
names(temp)=c("Name","Size")
temp$Type <- "Technology"

file$NewTechnology <- file$Technology
file$NewTechnology <- gsub(temp[which(temp$Size==1),"Name"][1], "Other",file$NewTechnology)
file$NewTechnology <- gsub(temp[which(temp$Size==1),"Name"][2], "Other",file$NewTechnology)
file$NewTechnology <- gsub(temp[which(temp$Size==1),"Name"][3], "Other",file$NewTechnology)
file$NewTechnology <- gsub(temp[which(temp$Size==1),"Name"][4], "Other",file$NewTechnology)

table_info <- file[,c("Study","Year","Journal","Blood/PBMC","Serum/Plasma","Sputum/Saliva","Urine","Other","Technology","NewTechnology","Multi-omics")]
table_info <- table_info[order(table_info$`Blood/PBMC`,table_info$`Serum/Plasma`,table_info$`Sputum/Saliva`,table_info$Urine,table_info$Other,table_info$NewTechnology,table_info$`Multi-omics`,table_info$Year,decreasing = TRUE),]
write.csv(table_info,file="Table-info.csv",quote = FALSE)

temp=as.data.frame(table(file$NewTechnology))
names(temp)=c("Name","Size")
temp$Type <- "Technology"
statistics_data <- rbind(statistics_data,temp)

temp=as.data.frame(table(file$`Multi-omics`))
names(temp)=c("Name","Size")
temp$Type <- "Multi-Omics"
temp$Name <- gsub(FALSE,"Proteomics",temp$Name)
temp$Name <- gsub(TRUE,"Multiomics",temp$Name)
statistics_data <- rbind(statistics_data,temp)

statistics_data$Size=as.numeric(statistics_data$Size)
write.csv(statistics_data,file="DonutPlots.csv",quote=FALSE)
```

